---
title: "Coding Categorical Variables"
author: 
  - "Ashley Holen"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 2/19/25
format: html
editor: visual
theme: spacelab
---

## Contrast Coding Categorical Predictors

Working with categorical predictors is an important aspect of building effective statistical models. Categorical predictors, also known as factors, represent qualitative data that can be organized into distinct groups, such as gender, geographic location, or product type.

In R, these predictors can be integrated into various modeling techniques, helping us identify relationships and make predictions based on different *levels* of these variables.

## UCLA Reference

There are many resources out there, but this is one I go back to again and again. I find it really easy to understand and to also use as a reference:

[**R LIBRARY CONTRAST CODING SYSTEMS FOR CATEGORICAL VARIABLES**](https://stats.oarc.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/)

## Set up your environment

### Load packages

```{r, echo = FALSE, message = FALSE}
library(lme4)
library(tidyverse)
```

### Load Data

Following the UCLA resource, we will use data frame called `hsb2` and focusing on the categorical variable `race`, which has four levels (1 = Hispanic, 2 = Asian, 3 = African American and 4 = Caucasian). We will use `write` as our dependent variable.

```{r}
# read in data from URL
hsb2 = read.table('https://stats.idre.ucla.edu/stat/data/hsb2.csv', header=T, sep=",")
```

Let's create `race.f`, which is the `race` data re-labeled and also turned into factors.

```{r}
hsb2$race.f = factor(hsb2$race, labels=c("Hispanic", "Asian", "African-Am", "Caucasian"))
```

### Inspect your data

Inspect your data below using `skimr`, `summary`, and `ggplot2`. You should also check out [this resource](https://www.openintro.org/data/index.php?data=hsb2), which describes the dataset.

```{r}
skim <- skimr::skim(hsb2)
View(skim)

summary(hsb2)
summary(hsb2$race)
summary(hsb2$race.f)

ggplot(hsb2, aes(x = race.f, y = write)) +
  geom_boxplot()

ggplot(hsb2, aes(x = race.f, y = write)) +
  geom_jitter(width = 0.2, height = 0) 

```

## Coding Schemes

There are several different coding schemes, and you can also generate your own scheme for your own purposes. For our class, we will focus on **Dummy**, **Simple**, **Deviation**, and **Helmert** Coding.

We've already seen dummy coding, but it is good to see it again.

## Class activity

Split into four groups. Explore your assigned coding scheme and answer the below five questions. Then, apply your coding scheme to the `hsb2` data and walk us through its interpretation.

## Dummy Coding (a.k.a. Treatment Coding)

1.  What comparisons are made with this coding scheme?\
    **Compares each level to the reference level, intercept being the cell mean of the reference group**

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?\
    **Dummy coding is the default scheme and probably the most common. It is useful for situations where you need to interpret the coefficients as the difference between a specific category and the reference group. Dummy coding is a statistical technique that uses a series of 0s and 1s to represent categorical data. It's a common way to convert categorical variables into numerical data so that models can process them.**

    **Dummy coding is preferred when you need to incorporate categorical variables (nominal or ordinal data) into a regression model because it allows you to represent each category as a binary variable.**

    **Categorical variables like gender (male/female), location (city, suburb), or marital status (single, married) are the best fit and make dummy coding useful and efficient.** 

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?

    **Dummy coding offers several key advantages: it simplifies interpretation, is compatible with statistical tests, reduces multicollinearity, and allows analysis of categorical data. By converting categorical variables into binary values (0/1), it enables easier comparison between categories, especially when using a reference level. This method is also widely supported by statistical software, making it user-friendly and efficient for modeling.**

    **It enhances model performance by allowing categorical data to be included in a model by converting it into numerical values. This provides more information to the model and potentially improves its accuracy in predicting the target variable.**

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?\
    **Limitations:**

    1.  **Multicollinearity Issues**

        -   **Dummy coding introduces multiple binary variables, one for each category except the reference category. If the reference category is not well chosen, multicollinearity can arise, affecting the stability and interpretability of regression coefficients.**

    2.  **Interpretation Challenges**

        -   **The coefficients in dummy coding represent differences from the reference category. If the reference category is not meaningful or if the interpretation of differences is not intuitive, it can make results harder to communicate.**

    3.  **Loss of Information in Ordinal Data**

        -   **When dummy coding is used for ordinal variables, it treats all categories as independent, ignoring the inherent order. This can lead to inefficient models and loss of meaningful relationships.**

    4.  **Increase in Model Complexity**

        -   **Dummy coding increases the number of predictor variables, especially when dealing with categorical variables with many levels. This can lead to overfitting in models with limited data.**

    **There are some situations where dummy coding can be misleading such as if one category is significantly underrepresented, comparisons against it may be unstable or misleading. Also, using dummy coding instead of ordinal encoding can discard valuable information about category ordering.**\

5.  How does this coding scheme affect the coefficients in regression models? What do the coefficients represent in the context of your chosen coding method?

    **In regression models, dummy coding affects the interpretation of coefficients by setting one category as the reference group while creating binary variables for the remaining categories. Each coefficient represents the difference in the dependent variable between a given category and the reference category, holding all other variables constant. The intercept represents the expected outcome for the reference group. Since all comparisons are relative, choosing a meaningful reference category is important to avoid misleading interpretations. If there are many categories, dummy coding can add complexity and may cause multicollinearity.**\

### Applying Dummy Coding

```{r}
# assign dummy coding

#the contrast matrix for categorical variable with four levels
contr.treatment(4)

#assigning the treatment contrasts to race.f
contrasts(hsb2$race.f) = contr.treatment(4)

# build and summarize your regression
dummy <- lm(write ~ race.f, data = hsb2)
summary(dummy)
```

Summarize these findings. What is your reference group, and how do the other groups compare with regards to their writing scores? What does the intercept value represent?\
\
**These findings demonstrate** **that the reference group is Hispanic and the mean scoring of 46.458, which is statistically significant since being above 0, but this can be disregarded. Next, Asian scores compared to Hispanic score is statistically significant, as the model found that the Asian score was 11.542 points higher on average than the Hispanic scores. It was found that the African American scores compared to the Hispanic scores is not statistically significant, as the African American scores are on average 1.742 points higher than the Hispanic scores. Lastly, it is statistically significant that the Caucasian scores are on average 7.597 points higher than the Hispanic scores. The model is comparing the means of "write" for level 1 to 2, 1 to 3, and 1 to 4.** **The R-squared (0.1071) means that about 10.7% of the variance in writing scores is explained by race.**

## Simple Coding

1.  What comparisons are made with this coding scheme?\

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?\

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?\

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?\

5.  How does this coding scheme affect the coefficients in regression models? What do the coefficients represent in the context of your chosen coding method?

### Applying Simple Coding

```{r}
# assign simple coding


# build and summarize your regression
simple <- lm(write ~ race.f, data = hsb2)
summary(simple)
```

Summarize these findings. What is your reference group, and how do the other groups compare with regards to their writing scores? What does the intercept value represent?\
\
(write here)\

## Deviation Coding

1.  What comparisons are made with this coding scheme?

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?

5.  How does this coding scheme affect the coefficients in regression models? What do the coefficients represent in the context of your chosen coding method?

### Applying Deviation Coding

```{r}
# assign deviation coding


# build and summarize your regression
dev <- lm(write ~ race.f, data = hsb2)
summary(dev)
```

Summarize these findings. Unlike dummy and simple coding, there is no traditional reference level in deviation coding. In this context, how do the groups compare regarding their writing scores? What does the intercept value represent?\
\
(write here)\

## Helmert Coding

1.  What comparisons are made with this coding scheme?

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?

5.  How does this coding scheme affect the coefficients in regression models? What do the coefficients represent in the context of your chosen coding method?

### Applying Helmert Coding

```{r}
# assign Helmert coding


# build and summarize your regression
helmert <- lm(write ~ race.f, data = hsb2)
summary(helmert)
```

Summarize these findings. In Helmert coding, each level is compared to the mean of subsequent levels, rather than having a traditional reference group. In this context, how do the groups compare regarding their writing scores? What does the intercept value represent?\
\
(write here)

## For next class: Releveling a predictor

We will do this together as a class on Friday. You don't need to work ahead, but if you wanted to explore releveling, here you go.

Releveling refers to the process of changing the reference level of a factor or categorical variable. This allows us to examine how different groups compare to a new reference category. By selecting a different reference level, we can gain new insights into the effects of categorical predictors in our analyses.

Releveling is applicable to Dummy, Simple, and Helmert coding, but NOT deviation coding. Why?

### To relevel:

You can use the below to define a reference level, or you can assign new values in your contrasts matrix.

```{r}
hsb2$race.f <- relevel(hsb2$race.f, ref = "Asian")
```
