---
title: "Assignment 2: Wrapping up regression"
author: 
  - "Ashley Holen"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 2/28/25
format: html
toc: true
editor: visual
theme: spacelab
---

# Assignment 2: Wrapping up linear and logistic regression\*

Assignment 2 covers linear and logistic regression models and includes many of the topics we have covered over the entire semester. You will be using `palmerpenguins` data, which includes a nice variety of continuous and categorical predictors.

To submit this assignment, render this file and save as a pdf. Upload the pdf to Canvas.

\*This assignment does not include mixed effects models, but you will be seeing that on another assignment.

## Load Packages

```{r, message = FALSE}
library(lme4)
library(tidyverse)
# install.packages("palmerpenguins") # install if needed
library(palmerpenguins)
library(viridis)
```

## Load Data

Load the `penguins` data and examine it below (use `summary()`, etc.)

```{r}
penguins <- penguins
summary(penguins)
colnames(penguins)
head(penguins)

ggplot(penguins, aes(species, body_mass_g, color = sex)) +
  geom_point()

ggplot(penguins, aes(x=species, y=body_mass_g, fill=species)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    ggtitle("boxplot with jitter") +
    xlab("")

ggplot(penguins, aes(flipper_length_mm, bill_length_mm, color = species, shape = sex)) + 
  geom_point()

max(penguins['year'])
min(penguins['year'])
```

## Question 1: Describe the data

What data is contained in this data set? Describe at least four variables (excluding `year`), including what they represent and their data type. Lastly, describe whether you think `year` would be a useful predictor in this data.

**species** - The species of penguin. there are 3 species present in this data set: Adeline, Chinstrap, and Gentoo. The data type of 'species' is factor.

**bill_length_mm -** The length of the penguin's bill in millimeters. It looks as though this is rounded to the nearest tenth. The data type of 'bill_length_mm' is double.

**body_mass_g -** The body mass in grams of the penguin. The data type of 'body_mass_g' is int.

**sex -** The sex of the penguin: male, female, or NA. The data type of 'sex' is factor.

When looking at the '**year**' variable, the usefulness of it being a predictor depends on the context of the analysis which we are conducting. For example, if we were to look at variables over time, then 'year' could become a predictor. But, if features of the penguins are relatively stable over the years, then 'year' would not be a useful predictor. If 'year' does not show a trend affecting other variables, it may serve a grouping purpose rather than predictive.

## Question 2: EDA

Explore your data visually. Create at least two visualizations that show the relationship between `flipper_length_mm` and its potential predictors.

```{r, warning = FALSE}

# flipper length and bill length 
ggplot(penguins, aes(flipper_length_mm, bill_length_mm, color = species, shape = sex)) + 
  geom_point() +
  geom_smooth(aes(group = species), method = "lm", se = FALSE)

# flipper length and species
ggplot(penguins, aes(x=species, y=flipper_length_mm, fill=species)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    ggtitle("boxplot with jitter") +
    xlab("")
```

## Question 3: Apply a linear regression

Fit a simple linear regression model predicting `flipper_length_mm` from `body_mass_g`. Interpret the slope and intercept.

```{r}
#linear regression model
lm.model <- lm(flipper_length_mm ~ body_mass_g, data=penguins)
summary(lm.model)
```

Interpret your model output in your own words below. Be sure to include a sentence explaining how `body_mass_g` impacts `flipper_length_mm` and whether or not the effect is significant.

**Answer:** Based on the output of this model, it can be concluded that first of all, the intercept is given as 136.7 mm, which is the predicted flipper length when the body mass is 0. This is shows high significance, but not a possibility within our data set so the significance is not applicable. Then, for every 1 g increase in body mass, on average flipper length increases by approximately 0.01528 mm, which is highly statistically significant and positively correlated relationship. The R squared value of 0.759 tells us that about 75.9% of variance in flipper length is described by body mass. Overall, as penguins get heavier, their flipper length tends to increase.

## Question 4: Apply a multiple linear regression

Fit a linear regression model predicting `flipper_length_mm` from both `body_mass_g` and `bill_length_mm`. Interpret the slopes and intercept.

```{r}
#multiple linear regresson model
lm.model2 <- lm(flipper_length_mm ~ body_mass_g + bill_length_mm, data=penguins)
summary(lm.model2)
```

Similar to Question 3, interpret the model output in your own words here:

**Answer:** Based on the output of this model, it can be concluded that first of all, the intercept is given as 122 mm. This means that in this model, predicted flipper length is 122 mm when both body mass and bill length are 0. Since logically, penguins do not have 0 body mass or bill length, this is shows high significance, but not a possibility within our data set so the significance is not applicable. Next, the body mass coefficient is 0.01305, which means that for every 1 g increase in body mass, flipper length increases by an average of 0.01305 mm. This is highly statistically significant and positively correlated relationship. Then, the bill length coefficient is 0.5492, which means that for every 1 mm increase in bill length, flipper length increases by 0.5492 mm. This is also a highly statistically significant and positively correlated relationship. Lastly, looking at the R squared of 0.7884, this model explains about 78.84% of variance in flipper length.

## Question 5: Include an interaction

Fit a simple linear regression model predicting `flipper_length_mm` from `body_mass_g,` `bill_length_mm`, and the interaction of the two. Interpret the slopes and intercept.

```{r}
#multiple linear regresson model with interaction
lm.model3 <- lm(flipper_length_mm ~ body_mass_g + bill_length_mm + body_mass_g:bill_length_mm, data=penguins)
summary(lm.model3)
```

Interpret the model output in your own words below. If there was a change in the pattern of significance, try to explain the logic below as well.

**Answer:** Based on the output of this model, it can be concluded that first of all, the intercept is given as 170.6 mm, which is the predicted flipper length when both body mass and bill length are 0. Since logically, penguins do not have 0 body mass or bill length, this is shows high significance, but not a possibility within our data set so the significance is not applicable. The intercept changed from 122 mm from the previous model. Next, the body mass coefficient is 0.0004364, which which means that for every 1 g increase in body mass, flipper length increases by an average of 0.0004364 mm when bill length is at 0. This is no longer statistically significant compared to the previous model. This is NOT highly statistically significant and is a positively correlated relationship. Then, the bill length coefficient is -0.5051, which which means that for every 1 mm increase in bill length, flipper length DECREASES by 0.5051 mm. This is also no longer statistically significant compared to the previous model. This is NOT highly statistically significant and is a NEGATIVELY correlated relationship, where in the previous model, it was positive. Lastly, the coefficient for the interaction of body mass and bill length is 0.0002707, which shows the joint influence that body mass and bill length have on flipper length. It means that as either variable body mass or bill length increase by 1 unit, the effect of the other increases flipper length on average by 0.0002707 mm. This is statistically significant and is a positively correlated relationship.

This model shows that body mass and bill length alone are not strong predictors, but their combined effect is significant. This model showed different significance due to its ability to capture more of a complex relationship between variables. Perhaps, penguins with both heavy mass and longer bills had very different flipper length than the model predicted from the individual variables. This shows that flipper length is not predicted from the sum of body mass and flipper length effects, but how they interact.

## Question 6: Compare models

Compare the models you built in Questions 4 and 5 using `anova()`.

```{r}
#anova comparison
anova(lm.model2, lm.model3)
```

-   Which is the better model? How do you know?

    -   **Answer:** The better model is Model 2, which is the multiple linear regression with an interaction (flipper_length_mm \~ body_mass_g + bill_length_mm + body_mass_g:bill_length_mm)). I know due to the significance codes demonstrating a p value showing a statistically significant comparison. The improvement between the two models is significant, showing that Model 2 is significantly better.

-   Is it possible to compare the models from Questions 3 and 5 using the same method? Why or why not?

    -   **Answer:** No, it is not possible to compare the linear model to the multiple linear regression with an interaction. This is due to their differing structures. anova is used for nested models, where one model should be the simplified version of the other. lm.model only contains body mass as a predictor, where lm.model3 includes bill length, body mass, and the interaction between the two. AIC would be a better choice in this case.

## Question 7: Categorical predictors

Build a linear model that includes a categorical predictor of your choice. It is fine to stick with dummy coding. Optional: apply a different coding scheme AND interpret the output correctly for +1 extra credit.

```{r}
#categorical linear model
summary(penguins$sex)
contrasts(penguins$sex)

lm.cat <- lm(flipper_length_mm ~ sex + body_mass_g, data=penguins)
summary(lm.cat)
```

-   What is the reference level of your categorical predictor?

    -   **Answer:** The reference level of 'sex', our categorical predictor, is female. It is represented as a 0 value, and the intercept through dummy coding.

-   What is your interpretation of this model output? Address all coefficients.

    -   **Answer:** Based on the output of this model, it can be concluded that first of all, the intercept is given as 134.6, which is the estimated flipper length when predictors sex is female and body mass is 0. Since body mass cannot be 0, this is shows high significance, but not a possibility within our data set so the significance is not applicable. Next, the 'sexmale' coefficient is -3.957, which represents the difference in flipper length between male and female penguins. Since female penguins are our reference level, this states that on average, male penguins have on average 3.957 SHORTER flippers than female penguins, when body mass is a constant. This is highly statistically significant and is a NEGATIVELY correlated relationship. Lastly, the body mass coefficient is 0.01624, which means that regardless of sex, for each 1 g increase in body mass, flipper length increases by an average of 0.01624 mm. This is highly statistically significant and is a positively correlated relationship.

## Question 8: Relevel your categorical variable

Relevel your categorical variable so that a **different** level becomes the reference. Then, run the same model you did in Question 7 and interpret the output.

Relevel:

```{r}
#releveling sex variable to male reference 
penguins$sex <- relevel(penguins$sex, ref = "male")
levels(penguins$sex)
```

Apply model from Question 7:

```{r}
#rerunning categorical linear model 
lm.cat <- lm(flipper_length_mm ~ sex + body_mass_g, data=penguins)
summary(lm.cat)
```

-   What is the new reference level of your categorical predictor?

    -   **Answer:** The new reference level of 'sex', our categorical predictor, is male. It is represented as a 0 value, and the intercept through dummy coding.

-   What is your interpretation of this new model output? Address all coefficients.

    -   **Answer:** Based on the output of this model, it can be concluded that first of all, the intercept is given as 130.7, which is the estimated flipper length when predictors sex is MALE and body mass is 0. Since body mass cannot be 0, this is shows high significance, but not a possibility within our data set so the significance is not applicable. Next, the 'sexfemale' coefficient is +3.957, which represents the difference in flipper length between male and female penguins. Since male penguins are our reference level, this states that on average, male penguins have on average 3.957 LONGER flippers than female penguins, when body mass is a constant. This is highly statistically significant and is a POSITIVELY correlated relationship. astly, the body mass coefficient is 0.01624, which means that regardless of sex, for each 1 g increase in body mass, flipper length increases by an average of 0.01624 mm. This is highly statistically significant and is a positively correlated relationship.

## Question 9: Apply a logistic regression

Apply a logistic regression. Include as many predictor variables as you'd like. Remember that your predicted outcome variable needs to be binary (or categorical with two levels).

\
Hint: You could use `sex` or create a binary variable of your own (e.g., Gentoo vs. non-Gentoo) to test your model.

```{r}
log.model <- glm(sex ~ body_mass_g, data = penguins, family = binomial)
summary(log.model)
exp(-0.0012398)
(1 - 0.998761) * 100
```

What are your key takeaways from this model?

**Answer:** First, this model outputs an intercept of 5.1625. This is the log odds of a penguin being male when body mass = 0 g. Since body mass cannot be 0, this is shows high significance, but not a possibility within our data set so the significance is not applicable. The coefficient for body mass is -0.0012398, which means that as body mass increases, the log odds of a penguin being male decreases. In simple terms, heavier penguins are more likely to be female, while lighter penguins are more likely to be male. This is highly statistically significant. Body mass is a strong predictor of sex in penguins. Then using exp, for every 1g increase in body mass, the odds of being male decrease by about 0.12%.

## Question 10: Synthesize the information

Imagine you’re a biologist studying penguin populations. Which predictors do you think are most important to measure or record in the field to predict flipper length? Why?

**Answer:** Based on the models that were run, it can be concluded that the most important predictors to predict flipper length are body mass and bill length, more specifically, their interaction. Looking at the visualizations I created, species also has an impact, due to the species seeming to have an overall range of possible flipper lengths. Also, sex has an impact. We found that female penguins are an average heavier, and body mass was found to impact flipper length.

## Bonus: Stepwise Regression

Perform stepwise regression to find the best model for an outcome of your choice. You will likely encounter an error – fixing that error and explaining your findings will earn you +1 extra credit. Show your work.

```{r}
#stepwise regression
m.all <- lm(flipper_length_mm ~ ., data = penguins)
#summary(m.all)

#stepwise_model <- step(m.all, direction = "both") #this gives an error
colSums(is.na(penguins)) #checking NAs per column 
nrow(penguins) #checking num of rows, 344 

penguins_clean <- na.omit(penguins) #cleaning the data of NA values
colSums(is.na(penguins_clean)) 
nrow(penguins_clean) # 333, we took out 11 rows

m.all <- lm(flipper_length_mm ~ ., data = penguins_clean)

stepwise_model <- step(m.all, direction = "both")
```

According to this stepwise regression, explain how the final model was selected.

**Answer:** The stepwise regression began with a model containing all variables available when predicting flipper length. flipper_length_mm \~ species + island + bill_length_mm + bill_depth_mm + body_mass_g + sex + year. The regression removes one variable at a time, and it began with sex. This lowered the AIC from 1060.16 to 1058.96. Then, the stepwise regression evaluated other variables to take out, but every option would have increased the AIC. Thus, the process stopped. Thus, all remaining predictors show to be important in predicting flipper length. This process removes unnecessary predictors while keeping essential ones.
